A Wait-free Queue with Poly-logarithmic Step Complexity

### 2022-12-22 podc_paper/related.tex

\paragraph{List-based Queues}
The MS-queue \cite{MS98} is a lock-free queue that has stood the test of time.
A version of it is
included in the standard Java Concurrency Package.  %java.util.concurrent.ConcurrentLinkedQueue
See the paper that introduced it for a survey of the early history of concurrent queues.
As mentioned above, the MS-queue suffers from the CAS retry problem because of contention at the front and back of the queue.
Thus, it is lock-free but not wait-free and has an amortized step complexity
of $\Theta(p)$ per operation.

@@

Kogan and Petrank~\cite{DBLP:conf/ppopp/KoganP11} 
added Herlihy's helping
technique~\cite{10.1145/114005.102808} to the MS-queue to obtain
a wait-free queue.
Ramalhete and Correia \cite{RC17} added a different helping mechanism.
In both cases, the helping mechanisms only add to the step complexity of the MS-queue.

@@

Morrison and Afek \cite{DBLP:conf/ppopp/MorrisonA13} also used a linked list of (circular) arrays.
To avoid the CAS retry problem, concurrent operations try to claim spots in an array using fetch\&increment instructions.
If livelock between enqueues and a racing dequeue prevent enqueues from claiming a spot,
the enqueues fall back on using a CAS to add a new array to the linked list, 
and the CAS retry problem reappears.
% I looked at the CRQ algorithm to implement each of the circular arrays.
% Here is an execution that causes livelock for 3 processes
% enqueue(A) reads t=0 and increments tail to 1
% enqueue(B) reads t=1 and increments tail to 2
% (*) dequeue reads h=0 and increments head to 1
% dequeue's CAS succeds changing array[0] to (1,R,bottom)
% dequeue reads t=2 at line 53
% enqueue(A) fails CAS on array[0]
% dequeue reads h=1 and increments head to 2
% dequeue's CAS succeeds changing array[1] to (1,R+1,bottom)
% enqueue(B) fails CAS on array[1]
% enqueue(A) reads t=2 and increments tail to 3
% enqueue(B) reads t=3 and increments tail to 4
% dequeue reads t=4 at line 53
% go back to (*) and repeat this sequence
Similarly,
other queues \cite{Nik19,10.1145/3490148.3538572,DBLP:conf/ppopp/YangM16} used the fast-path slow-path methodology of Kogan and Petrank~\cite{10.1145/2370036.2145835}.
The fast path has good performance and the slow path guarantees
termination.
In worst-case executions, where processes must execute the slow path,
these implementations also take $\Omega(p)$ steps per operation, 
due either to the CAS retry problem or helping mechanisms.

>>>>

\paragraph{List-based Queues}
The MS-queue \cite{MS98} is a lock-free queue that has stood the test of time.
A version of it is
included in the standard Java Concurrency Package.  %java.util.concurrent.ConcurrentLinkedQueue
See the paper that introduced it for a survey of the early history of concurrent queues.
As mentioned above, the MS-queue suffers from the CAS retry problem because of contention at the front and back of the queue.
Thus, it is lock-free but not wait-free and has an amortized \here{Maybe we can say worst-case. I see that you have used amortized step complexity consistently everywhere but maybe it is fair to say that these papers were not designed to improve the worst-case amortized step complexity.} step complexity
of $\Theta(p)$ per operation.

@@

Kogan and Petrank~\cite{DBLP:conf/ppopp/KoganP11} 
added Herlihy's helping
technique~\cite{10.1145/114005.102808} to the MS-queue to obtain
a wait-free queue.
Ramalhete and Correia \cite{RC17} added a different helping mechanism.
In both cases, the helping mechanisms only add to the step complexity of the MS-queue. \here{I get what this sentence means, but from a high overview i think it is inaccurate. Since the step complexity of MS-queue is infinite, so nothing can be added to it and a wait-free queue is always better than a lock-free queue in the worst-case.}

@@

Morrison and Afek \cite{DBLP:conf/ppopp/MorrisonA13} also used a linked list of (circular) arrays.
To avoid the CAS retry problem, concurrent operations try to claim spots in an array using fetch\&increment instructions.
If livelock between enqueues and a racing dequeue prevent enqueues from claiming a spot,
the enqueues fall back on using a CAS to add a new array to the linked list, 
and the CAS retry problem reappears.
% I looked at the CRQ algorithm to implement each of the circular arrays.
% Here is an execution that causes livelock for 3 processes
% enqueue(A) reads t=0 and increments tail to 1
% enqueue(B) reads t=1 and increments tail to 2
% (*) dequeue reads h=0 and increments head to 1
% dequeue's CAS succeds changing array[0] to (1,R,bottom)
% dequeue reads t=2 at line 53
% enqueue(A) fails CAS on array[0]
% dequeue reads h=1 and increments head to 2
% dequeue's CAS succeeds changing array[1] to (1,R+1,bottom)
% enqueue(B) fails CAS on array[1]
% enqueue(A) reads t=2 and increments tail to 3
% enqueue(B) reads t=3 and increments tail to 4
% dequeue reads t=4 at line 53
% go back to (*) and repeat this sequence
\here{Is it ok for a reader to talk about A and then start a sentence about similarity to B? My suggestion which is longer: Kogan and Petrank~\cite{10.1145/2370036.2145835} proposed the fast-path slow-path methodology which is similar to Morrison and Afek \cite{DBLP:conf/ppopp/MorrisonA13}. Other queues \cite{Nik19,10.1145/3490148.3538572,DBLP:conf/ppopp/YangM16} have used the methodology.} Similarly,
other queues \cite{Nik19,10.1145/3490148.3538572,DBLP:conf/ppopp/YangM16} used the fast-path slow-path methodology of Kogan and Petrank~\cite{10.1145/2370036.2145835}.
The fast path has good performance and the slow path guarantees
termination.
In worst-case executions, where processes must execute the slow path,
these implementations also take $\Omega(p)$ steps per operation, 
due either to the CAS retry problem or helping mechanisms.
